---
title: "Spotify Descriptive Statistics"
author: "Matthew, Sam, Andrew"
format: pdf
editor: visual
---

# STAT 3120

Statistical Analysis of Spotify Dataset.

## Setup

```{r}
#install.packages(c("readxl", "dplyr", "ggplot2", "janitor", "epitools", "gt", "car"))
```

```{r}
#| message: false
#| warning: false
library(readxl)
library(dplyr)
library(ggplot2)
library(janitor)
library(epitools)
library(gt)
library(car)
```

```{r}
# Export Dataset, substitute "path/to/file" with actual path
sp <- read_excel("Data/spotify_stat3120.xls")
# View(sp) # To view dataset in R
```

# Descriptive Statistics and Cleaning

## Danceability (Quantitative)

```{r}
dance <- sp$danceability
summary(dance) # min 0.166, Q1 0.511, med 0.636, mea 0.629, Q3 0.754, max 0.954
sd(dance)      # 0.159
```

## Energy (Quantitative)

```{r}
energy <- sp$energy
summary(energy) # min 0.109, Q1 0.643, med 0.823, mea 0.778, Q3 0.935, max 0.998
sd(energy)      # 0.178
```

## Loudness (Quantitative, cleaning required)

```{r}
clean_sp <- filter(sp, loudness < 0) # remove values greater than 0
loud <- clean_sp$loudness
var(loud)  # 8.166
sd(loud)   # 2.858
summary(loud) # min -17.912, Q1 -8.063, med -6.126, mea -6.471, Q3 -4.737, max -0.344
remaining_loud <- (length(sp$loudness) - length(loud)) # 5 values were removed
remaining_loud
```

## Tempo (Quantitative)

```{r}
tempo <- sp$tempo
summary(tempo) # min 81.99, Q1 129.98, med 145.00, mea 147.22, Q3 160.18, max 220.04
sd(tempo)      # 23.244
```

## Qualitative Variables

```{r}
table(sp$genre)
# Dark Trap 43, DnB 38, Emo 20, Hardstyle 48, HipHop 38, Pop 2, Psytrance 50, Rap 17
# RnB 20, Techhouse 44, Techno 32, Trance 40, Trap 38, Trap M 34, Underground 79.
```

## Creating DanceCat by separating Danceability by the mean

```{r}
sp$dancecat <- ifelse(sp$danceability > .629, "Danceable", "Not Danceable")
```

## Removing Pop because of low count of 2

```{r}
clean_sp <- sp %>% filter(genre != "Pop")
```

## Contingency table and expected values

```{r}
# New contingency table
tab <- table(clean_sp$dancecat, clean_sp$genre)
tab
```

```{r}
# Expected values
expected <- chisq.test(tab)$expected
expected
```

# Graphs

## Boxplot with histogram for Danceability

```{r}
layout(mat = matrix(c(1,2),2,1, byrow=TRUE), height = c(1,8))
par(mar=c(0, 3.1, 1.1, 2.1))
boxplot(dance, horizontal=TRUE, ylim=c(0,1), xaxt="n",
        col=rgb(0.8,0.8,0,0.5), frame=F)
par(mar=c(4, 5, 1.1, 2.1))
hist(dance, breaks=40, col=rgb(0.2,0.8,0.5,0.5), border=F,
     main="", xlab="Danceability", ylab="Frequency", xlim=c(0,1))
```

## Boxplot with histogram for Energy

```{r}
layout(mat = matrix(c(1,2),2,1, byrow=TRUE), height = c(1,8))
par(mar=c(0, 3.1, 1.1, 2.1))
boxplot(energy, horizontal=TRUE, ylim=c(0,1), xaxt="n",
        col=rgb(0.2, 0.4, 0.8, 0.5), frame=F)
par(mar=c(4, 5, 1.1, 2.1))
hist(energy, breaks=40, col=rgb(0.2, 0.4, 0.8, 0.8), border=F,
     main="", xlab="Energy", ylab="Frequency", xlim=c(0,1))
```

## Boxplot with histogram for Loudness

```{r}
layout(mat = matrix(c(1,2),2,1, byrow=TRUE), height = c(1,8))
par(mar=c(0, 3.1, 1.1, 2.1))
boxplot(loud, horizontal=TRUE, ylim=c(-30,0), xaxt="n",
        col=rgb(0.2, 0.7, 0.4, 0.5), frame=F)
par(mar=c(4, 5, 1.1, 2.1))
hist(loud, breaks=40, col=rgb(0.1, 0.5, 0.3, 0.8), border=F,
     main="", xlab="Loudness", ylab="Frequency", xlim=c(-30,0))
```

## Boxplot with histogram for Tempo

```{r}
layout(mat = matrix(c(1,2),2,1, byrow=TRUE), height = c(1,8))
par(mar=c(0, 3.1, 1.1, 2.1))
boxplot(tempo, horizontal=TRUE, ylim=c(50,250), xaxt="n",
        col=rgb(0.6, 0.3, 0.8, 0.5), frame=F)
par(mar=c(4, 5, 1.1, 2.1))
hist(tempo, breaks=40, col=rgb(0.7, 0.2, 0.6, 0.8), border=F,
     main="", xlab="Tempo", ylab="Frequency", xlim=c(50,250))
```

## Bar plot for Genre

```{r}
ggplot(sp, aes(x = genre)) + geom_bar()
```

```{r}
# without pop
ggplot(clean_sp, aes(x = genre)) + geom_bar()
```

## Stacked Proportional bar chart (genre vs dance)

```{r}
ggplot(clean_sp, aes(x = genre, fill = dancecat)) +
  geom_bar(position = "fill") +
  labs(y = "Proportion")
```

# Statistical Tests

## Test 1: Multiple Linear Regression Analysis

```{r}
# Regression Test
model <- lm(danceability ~ loudness + tempo, data = sp)
summary(model) # df=540, f-stat=25.09, p=3.803e-11
```

### Checking conditions

```{r}
vif(model)        # Multicollinearity check, passed bc close to 1.
```

```{r}
plot(model, which = 1) # Residual plot, passed graph looks decently centered around the middle.
```

```{r}
plot(model, which = 2) # Q-Q plot, passed no jarring signs.
```

## Test 2: Chi-Square Test of Independence

```{r}
# Verifying expected counts
all(expected >= 5) # true, satisfied
```

```{r}
# Chi Square Test-of-Independence
chi_result <- chisq.test(tab)
chi_result # x-squared=186.87, df=13, p<2.2e-16
```
